---
title: "Spatial Models in Stan: Intrinsic Auto-Regressive Models for Areal Data"
author: "Mitzi Morris"
output: 
  html_document:
    keep_md: false
---
<style type="text/css">
.table {

    width: 40%;

}
</style>

When data has a spatio-temporal structure and when observations from
neighboring regions exhibit higher correlation than observations between
distant regions,  this correlation can be accounted for using the class
of spatial models called "CAR" models (Conditional Auto-Regressive)
introduced by Besag (Besag 1974).
Intrinsic Auto-Regressive (IAR) models are a subclass of CAR models.
The Besag York Mollié (BYM) model is a lognormal Poisson model which
includes both an IAR component for spatial smoothing and an ordinary
random-effects component for non-spatial heterogeneity.
This case study covers how to efficiently code these models in Stan.

All models and data files are available in the Stan example-models GitHub repo for Stan case studies:
[car-iar-poisson](https://github.com/stan-dev/example-models/tree/master/knitr/car-iar-poisson).
All commands should be run from the directory `stan-dev/example-models/knitr/car-iar-poisson`.

## Formulations of CAR and IAR models

Given a set of observations taken at $n$ different areal units
of a region with a number of dimensions $D$
(for spatio-temporal data, this number is between 1 and 4
as there are 1-3 spatial dimensions and 1 time dimension),
spatial interactions between regions $n_i$ and $n_j$ can be modelled conditionally
as a spatial random variable $\mathbf{\phi}$, which is an $n$-length vector
$\mathbf{\phi} = ({\phi}_1, \ldots, {\phi}_n)^T$.

We represent spatial relationship between the $n$ areal units
as an adjacency matrix $W$ with dimensions $n \times n$
where entries $w_{ij}$ and $w_{ji}$ are positive when regions ${n_i}$ and ${n_j}$ are neighbors
and zero otherwise.
The _neighbor_ relationship $i \sim j$ is defined in terms of this matrix:
the neighbors of region $i$ are those regions who have non-zero entries in row or column $i$.
This encoding defines a lattice structure over the $n$ areal units.

Besag (1974) uses results from the physics of lattice systems of particles
and the Hammersley-Clifford theorem provide an equivalence between
a local specification of the conditional distribution of each particle
given its neighboring particles and the global specification
of the joint distribution of all particles.
This specification of the joint distribution via the local specification
of the conditional distributions of the individual variables
is a Markov random field (MRF) specification.
The conditional distribution for each ${\phi}_i$ is specified in terms of a mean
and precision parameter $\tau$ as:

$$ p \left( { \phi }_i \, \vert\, {\phi}_j \, j \neq i, {{\tau}_i}^{-1} \right)
= \mathit{N} \left( \alpha \sum_{i \sim j} {w}_{ij} {\phi}_j,\tau_i^{-1} \right), i,j = 1, \ldots, n $$

The corresponding joint distribution can be uniquely determined from
the set of full conditional distributions by
introducing a fixed point from the support of $p$
and then using Brook’s Lemma to factor the set of conditional distributions
into a joint distribution which is determined up to a proportionality constant
(see Banerjee, Carlin, and Gelfand, 2004, sec. 3.2):

$$ \mathbf{\phi} \sim \mathit{N} \left(\mathbf{0}, \left[D_{\tau}(I - \alpha B)\right]^{-1} \right) $$

where

- $B$ is an $n \times n$ matrix where entries $\{i,i\}$ are zero and the off-diagonal elements
describe the spatial proximity of regions $n_i$ and $n_j$.
- $I$ is an $n \times n$ identity matrix
- $D_{\tau} = \tau D$ where D is an $n \times n$ diagonal matrix.

The construction of the spatial proximity matrix $B$ determines the class of CAR model structure.
The model is an Intrinsic Auto-Regressive (IAR) model when:

- $\alpha = 1$
- $D$ is an $n \times n$ diagonal matrix where $d_{ii}$ = the number of neighbors for region $n_i$
- $W$ is the adjacency matrix where $w_{ii} = 0, w_{ij} = 1$ if $i$ is a neighbor of $j$, and $w_{ij}=0$ otherwise
- $B = D^{-1} W$, i.e., $B$ is the scaled adjacency matrix

The corresponding conditional distribution specification is:

$$ p \left( { \phi }_i \, \vert\, {\phi}_j \, j \neq i, {{\tau}_i}^{-1} \right)
= \mathit{N} \left( \frac{\sum_{i \sim j} {\phi}_{i}}{d_{ii}}, \frac{1}{d_{ii} {\tau}_i} \right)$$

where $d_{ii}$ is the number of neighbors for region $n_i$.
Intuitively, the individual spatial random variable ${\phi}_j$ for region $n_j$ with neighbors $N(i)$
is normally distributed with a mean equal to the average of its neighbors.
Its variance decreases as the number of neighbors increases.

The joint distribution simplifies to the _pairwise difference_ formulation:

$$\phi \sim N(0, [\tau \, (D - W)]^{-1}).$$

which rewrites to:

$$ p(\phi | \tau) \propto \frac{1}{{\tau}^\frac{n}{2}} \exp \left\{ {- \frac{1}{2\tau}} \sum_{i \sim j}{({\phi}_i - {\phi}_j)}^2 \right\} $$

The above conditions for the IAR model produce an improper distribution
because setting $\alpha = 1$ creates a singular matrix $(D - W)$, see Besag and Kooperberg 1995.
Furthermore, the pairwise distance formulation of the joint distribution is non-centered;
adding any constant to all of the elements of $\phi$ leaves the joint distribution unchanged.
To correct this, we add the constraint $\sum_{i} {\phi}_i = 0$.

While this IAR model is non-generating in that it cannot be used as a model for the data,
it can be used as a prior as part of a hierarchical model, which is the role it plays in
the BYM model.


## Multi-level Poisson GLMs: the Besag York Mollié (BYM) Model

Adding a CAR spatially structured error term to a multi-level GLM
provides spatial smoothing of the resulting estimates.
For count data, e.g. incidents of disease or accidents,
Poisson CAR models are available via R, BUGS, and JAGS
as well as INLA (Integrated Nested Laplace Approximation)
which is a fast alternative to MCMC,
(INLA trades speed and scalability for accuracy,
per the "no free lunch" principle).

The lognormal Poisson model proposed in Besag York Mollié 1991
is used for count data in biostatistics and epidemiology.
It includes both an IAR component for spatial smoothing and an
ordinary random-effects component for non-spatial heterogeneity.
Banerjee Carlin and Gelfand 2003, section 5.4, presents the details of
this model and its difficulties, together with a WinBUGS implementation
which they use to fit the Scottish lip cancer dataset from Clayton and Kaldor 1987.
Using the notation of Banerjee et al., the Besag York Mollié model is:

$$
Y_i \vert \psi_i \sim Poisson ( E_i \, e^{\psi_i}),
$$

for $i \in 1:N$, where 

$$
\psi = x \beta + \theta + \phi
$$

where

- $x$ is the matrix of explanatory spatial covariates, so that $x_i$ is
the vector of covariates for areal unit $i$.  The coefficients $\beta$ are
often called "fixed effects."

- $\theta$ is an ordinary random-effects components for non-spatial heterogeneity.

- $\phi$ is an IAR spatial component.

The pairwise difference formulation of the IAR spatial component $\phi$ is non-centered,
thus models with include both an IAR spatial effects component and an intercept term are non-identifiable.
Adding the constraint that $\phi$ must sum to zero centers it, allowing the model to
fit both the fixed-effect intercept $\beta\_0$ as well as $\phi$ and $\theta$.


### A Stan Implementation of the BYM Model

A Stan model which implements the BYM model for univariate data plus offset is in the file
[bym_predictor_plus_offset.stan](https://github.com/stan-dev/example-models/tree/master/knitr/car-iar-poisson/bym_predictor_plus_offset.stan).

```{r print-stan-bym, comment='', echo = FALSE}
cat(readLines('bym_predictor_plus_offset.stan'), sep = '\n')
```

In the next sections we discuss the data structures and coding idioms used
to insure good fit and efficient performance.

#### Encoding the Adjacency Matrix

The information needed to compute the IAR spatial variance component of the model is:

- $D$ - the $n \times n$ diagonal matrix where $d_{ii}$ = the number of neighbors for region $n_i$
- $W$ - the adjacency matrix where $w_{ii} = 0, w_{ij} = 1$ if $i$ is a neighbor of $j$, and $w_{ij}=0$ otherwise

A straightforward encoding of the adjacency matrix is as a
square matrix with dimensions equal to the number of areal units.
Depending on the regularity or irregularity of areal boundaries, the number of neighbors
for each region may vary.  In a completely regular grid structure each region will have
4 neighbors with which it shares an edge and another 4 regions diagonally adjacent,
thus in any row or column, the number of non-zero entries is 4 or 8.
The size of the adjacency matrix grows quadratically, while the number of non-zero entries
grows linearly, thus the matrix quickly becomes very sparse.
Representing a sparse matrix as a full matrix is expensive both
in terms of storage and computation; 
an adjacency matrix for $n$ areal units requires $n^2$ units of memory, and
walking over all entries in this matrix requires many needless lookups.

The information needed from $D$ for the IAR spatial component is
an array or vector of length $n$ where each entry $n_i$ contains the 
number of neighbors for region $n_i$.

The information needed from the adjacency matrix is a list of the {i,j}
coordinates of the non-zero entries in this matrix.
Since these entries all have value $1$, we don't need to store this value itself.
Since the adjacency matrix is symmetric, we only need store the entries in the
upper or lower triangular matrix.

The data block of the Stan model declares these structures as:

```
  int<lower=0> N;                 // num regions
  int<lower=0> N_edges;           // num edges, (undirected)
  int<lower=1, upper=N> node1[N_edges];  // node1[i] adjacent to node2[i]
  int<lower=1, upper=N> node2[N_edges];  // node1[i] adjacent to node2[i]
```

To wrangle the data into this format, we have written two R helper functions:

- [mungeCARdata4stan.R](https://github.com/stan-dev/example-models/tree/master/knitr/car-iar-poisson/mungeCARdata4stan.R)
- [nbdata4stan.R](https://github.com/stan-dev/example-models/tree/master/knitr/car-iar-poisson/nbdata4stan.R)

The function `mungeCARdata4stan` is used to translate spatial data from the format required by WinBUGS `car.normal` function.
It takes two inputs, `adj` which is a list of all neighbors by site id
and `num` which is a list of the number of neighbors for each site.
It returns a list containing the input data objects `N`, `N_edges`, `node1`, and `node2`.
The function `nbdata4stan` is used to translate "nb" objects generated by R's
[spdep](https://cran.r-project.org/web/packages/spdep) package.
It takes an `nb` object as input and returns a list 
containing the input data objects `N`, `N_edges`, `node1`, and `node2`.

For both of these functions, the conditional on the innermost loop
restricts the edges recorded in the parallel arrays `node1` and `node2`
to entries in the upper triangular part of the adjacency matrix.

#### Computing the Poisson-IAR Model

The model block specifies the vectorized Poisson regression, working on the log scale:
```
  y ~ poisson_log(log_E + beta0 + beta1 * x + theta + phi);
```

We use the non-centered parameterization for both random effects components `theta` and `phi`.

The values for the priors are taken from a model on Brad Carlin's website,
also discussed in Banerjee, Carlin, and Gelfand, 2004, chapter 5:
```
  beta0 ~ normal(0, inv(sqrt(1e-5)));
  beta1 ~ normal(0, inv(sqrt(1e-5)));
  theta_std ~ normal(0, 1);
  tau_theta ~ gamma(3.2761, 1.81);
  tau_phi ~ gamma(1, 1);
```

Computation of the IAR spatial component $\phi$ is distributed across the
transformed parameters block and the model block.

$\phi$ is constrained $\sum_{i} {\phi}_i = 0$.
To meet the sum-to-zero constraint we declare the parameter vector `phi_std_raw` of length $N-1$.
In the transformed parameters block we declare the $N$-length vector `phi` and ensure that it is
centered on zero by defining the N-th element as the negation of the sum of the elements 1 through N − 1.

```
parameters {
  ...
  vector[N - 1] phi_std_raw; // raw, standardized spatial effects
  ...
}
transformed parameters {
  ...
  phi[1:(N - 1)] = phi_std_raw;
  phi[N] = -sum(phi_std_raw);
  phi = phi * sigma_phi;
  ...
}
```

Computation of the pairwise difference is done in the model block.
From above, $\phi$ is a multivariate normal, with variance $[\tau (D - W)]^{-1}$, which rewrites to:

$$ p(\phi | \tau) \propto \frac{1}{{\tau}^\frac{n}{2}} \exp \left\{ {- \frac{1}{2\tau}} \sum_{i \sim j}{({\phi}_i - {\phi}_j)}^2 \right\} $$

The corresponding Stan statement is:
```
  target += -dot_self(phi[node1] - phi[node2]);
```

The expression `phi[node1] - phi[node2]` uses Stan's multiple indexing feature to specify
an array of integer indexes, thus this is a vectorized subtraction operation, the result of which is a vector.
The `dot_self` function which returns the dot product of this vector with itself, thus computing sum of the squares.
Because Stan is computing $\phi$ on the unit scale, $\tau$ is 1, and drops out.
The $\frac{1}{2}$ drops out because of symmetry:  by using only the upper triangular entries in the
adjacency matrix, we want to count them twice.
Thus by clever construction of the prior, the data, vectorization, and the fact that Stan computes
densities on the unit scale only up to a proportional constant, we can efficiently compute
the IAR spatial component $\phi$.


#### Fitting the Model to the Scotland Lip Cancer Dataset

To check this model, we run it over the version of the Scotland Lip Cancer dataset
from [Brad Carlin's software page](http://www.biostat.umn.edu/~brad/data2.html),
in file "Lipsbrad.odc".
This file contains both the data and the program and inits for the WinBUGS model
presented in Banerjee, Carlin, and Gelfand, 2004, chapter 5.
Following the Carlin model, where covariate `aff` is divided by 10, we multiply the `x` covariate by 0.1,
as multiplication is more efficient than division.
A Stan model which matches the Carlin model is in the file
[bym_carlin_scotland.stan](https://github.com/stan-dev/example-models/tree/master/knitr/car-iar-poisson/bym_carlin_scotland.stan).
It is exactly the same as the above model, save for the Poisson regression which is:
```
  y ~ poisson_log(log_E + beta0 + 0.1 * beta1 * x + theta + phi);
```

We run this program using the Rstan interface via the R script:
[fit_scot_stan.R](https://github.com/stan-dev/example-models/tree/master/knitr/car-iar-poisson/fit_scot_stan.R)
```{r fit-scot-stan, echo = FALSE, comment = NA, message = FALSE, results = "hide", warning = FALSE}
source("fit_scot_stan.R")
```
After running this script, we print a subset of the fitted parameters to check that the sampler was
able to fit the data; all Rhat values should be very close to 1.0, and the number of effective samples
should be in the thousands.


```{r print-fit-scot-stan, echo = FALSE, comment = NA, message = FALSE, warning = FALSE}
print(bym_stanfit, digits=4,
pars=c("lp__", "beta0", "beta1", "tau_theta", "tau_phi", "sigma_theta", "sigma_phi", "psi", "eta[1]", "eta[2]"),
probs=c(0.025, 0.975))
```

### A WinBUGS Implementation of the BYM Model

The program and data for the WinBUGS model, are available from
[Brad Carlin's software page](http://www.biostat.umn.edu/~brad/data2.html),
in file "Lipsbrad.odc".

The WinBUGS model is in file [carlin_bym.txt](https://github.com/stan-dev/example-models/tree/master/knitr/car-iar-poisson/carlin_bym.txt).

```{r print-bugs-scot-bym, comment='', echo = FALSE}
cat(readLines('carlin_bym.txt'), sep = '\n')
```

The data file [carlin_data.R](https://github.com/stan-dev/example-models/tree/master/knitr/car-iar-poisson/carlin_data.R)
contains the following objects:

- `O`: the observed lip cancer case counts on a per-county basis
- `aff`: an area-specific continuous covariate that represents the proportion
of the population employed in agriculture, fishing, or forestry (AFF)
- `E`: the an expected number of cases, used as an offset,
- `adj`: a list of region ids for adjacent regions
- `num`: a list of the number of neighbors for each region

Data items `adj` and `num` are used as input to the WinBUGS function `car.normal` which
computes the values of the elements for the zero-centered spatial component $\phi$.

In the WinBUGS program, the formula for the Poisson regression is:

```
O[i] ~ dpois(mu[i])
log(mu[i]) <- log(E[i]) + beta0 + beta1*aff[i]/10 + phi[i] + theta[i]
```
(Note that the covariate term `aff` is divided by 10).

To compare this model to the Stan model, 
we include an R script to run the WinBUGS model in OpenBUGS on the same data:
[fit_scot_bugs.R](https://github.com/stan-dev/example-models/tree/master/knitr/car-iar-poisson/fit_scot_bugs.R)
After running the model, we use the function `rstan::monitor` to compare the Stan fit with OpenBUGS fit of
the WinBUGS model.
The WinBUGS model uses the suffix ".h" for the heterogeneous random effects component which is "theta"
in the Stan model, and it uses the suffix ".c" for the IAR spatial effect component which is "phi" in the Stan model.
The WinBUGS program "alpha" corresponds to the Stan quantity "psi" and the WinBUGS "xi"
corresponds to the Stan quantities "eta".

```
                mean  se_mean      sd      2.5%        25%        50%        75%     97.5% n_eff   Rhat
beta0      -0.292821 0.004952 0.16362  -0.62310  -0.402500  -0.291000 -1.829e-01   0.02528  1092 1.0024
beta1       0.430032 0.005346 0.15983   0.11229   0.324500   0.431900  5.372e-01   0.73861   894 1.0032
tau.h       4.630522 0.025351 1.28887   2.50900   3.712500   4.490000  5.402e+00   7.52117  2585 1.0055
tau.c       2.517394 0.026565 1.00376   1.07000   1.790000   2.348000  3.057e+00   4.92312  1428 1.0014
sd.h        0.428034 0.001392 0.06218   0.31800   0.384600   0.424000  4.663e-01   0.56290  1996 1.0064
sd.c        0.576708 0.002922 0.09117   0.40339   0.514800   0.574900  6.357e-01   0.76640   974 1.0058
alpha       0.572579 0.001709 0.05567   0.45330   0.537900   0.576200  6.115e-01   0.67130  1061 1.0103
xi[1]       1.232006 0.004781 0.34157   0.53658   1.009000   1.239000  1.469e+00   1.88002  5104 1.0007
xi[2]       1.066722 0.004442 0.20843   0.66020   0.925175   1.068000  1.208e+00   1.47402  2202 1.0011
```

These estimates are not within `se_mean` of one another, although
they do get similar estimates for the slope and intercept terms of the poisson regression
as well as the quantities `eta` and `xi`.
As noted above, the Stan models use the non-centered parameterization for both `theta` and `phi`.
This could also be due to the difference between the Gibbs sampler and the NUTS HMC sampler.

## From 56 Counties in Scotland to 709 Census Tracts in Brooklyn

To demonstrate the scalability of using Stan to compute a spatial IAR component,
we analyze the Brooklyn subset of the dataset from
[Small-area spatiotemporal analysis of pedestrian and bicyclist injuries in New York City](https://www.ncbi.nlm.nih.gov/pubmed/25643104)
which was compiled from all reported traffic accidents involving a car and either a pedestrian or bicyclist
in New York City in 2001, localized to the census tract level.

The traffic accident data is in the file `nyc_ped_subset.data.R`.
It contains a list of census tract IDs,
the count of events per tract (`y`),
and the rate-adjusted population per tract (`x`).

```{r nyc-data-xy-plot, echo = FALSE, comment = NA, message = FALSE, warning = FALSE}
source("nyc_ped_subset.data.R");
y = events_all_2001[all_tractIDs %in% bklyn_tractIDs];
x = pop_adj_2001[all_tractIDs %in% bklyn_tractIDs];

plot(x,y,xlab="rate-adjusted population (scaled)",ylab="observed events");
```

The Stan program is in the file
[bym_predictor_only.stan](https://github.com/stan-dev/example-models/tree/master/knitr/car-iar-poisson/bym_predictor_only.stan).
This program implements the BYM model for univariate data without an offset term.

Spatial information is in a set of files in directory `nycTracts10`.
The spatial information for the census tracts is obtained via the R `maptools` and `spdep` packages.
We use these packages to
create an `nb` object which is a list of all neighbors for each census tract.
Each list entry is itself a list containing the relative index of the neighboring regions.
We have written an R helper function `nbdata4stan` that take an `nb` object as input
and returns a list containing the input data objects `N`, `N_edges, `node1`, and `node2`.

We fit this model using only the Brooklyn census tracts, so that all areal units have at least one neighbor.
There are 709 census tracts, more than 10x's the number of counties in the Scotland lip cancer dataset.
We assemble the Brooklyn data and fit the model Rstan interface via the R script:
[fit_brooklyn.R](https://github.com/stan-dev/example-models/tree/master/knitr/car-iar-poisson/fit_brooklyn.R)
```{r fit-brooklyn, echo = FALSE, comment = NA, message = FALSE, results = "hide", warning = FALSE}
source("fit_brooklyn.R")
```

This script uses R's `proc.time()` function to measure how long this model takes to fit the data.
On a MacBook Pro with 8G of memory and 2 processors, the reported times in seconds are:
```
   user  system elapsed 
 21.097   1.632 200.232 
```

The Rhat and Neff numbers indicates that the model has fit the data:
```{r print-brooklyn-fit, echo = FALSE, comment = NA, message = FALSE, warning = FALSE}
print(bk_stanfit,
      pars=c("lp__", "beta0", "beta1", "tau_theta", "tau_phi", "sigma_theta", "sigma_phi", "psi", "eta[1]", "eta[2]", "eta[3]", "eta[4]", "eta[709]"),
      probs=c(0.025, 0.5, 0.975));
```

We can compare these results to using a simple Poisson GLM to fit this data:
```{r brooklyn-poisson-glm-fit, echo = FALSE, comment = NA, message = FALSE, warning = FALSE}
r_glm_poisson = glm(y~x,family="poisson")
print(r_glm_poisson)
```

We use `ggplot2` and related packages to plot
the Brooklyn census tracts colored by event counts, scaled population,
and the generated quantity "mu" which recovers
the number of events based on the data and the samples.
Note that some Brooklyn census tracts are not in the study.

```{r setup-plot-data, include = FALSE, echo = FALSE, comment = NA, message = FALSE, results = "hide" }
library(gpclib);
library(maptools);  
library(spdep);
gpclibPermit()
library(dplyr) 
library(ggplot2) 
library(gridExtra)

mu_names = paste("mu[", c(1:709), "]")
brooklyn_samples = as.data.frame(bk_stanfit, pars=mu_names)
brooklyn_means = apply(brooklyn_samples, 2, mean)

incpt = coef(r_glm_poisson)["(Intercept)"]
slope = coef(r_glm_poisson)["x"]
r_glm_poisson_ys = exp(x * slope + incpt);

# prepare dfs for ggplot2
bklyn_events = events_all_2001[bklyn_tracts]
bklyn_pop = pop_adj_2001[bklyn_tracts]
df1 = data.frame(bklyn_tractIDs, bklyn_events, bklyn_pop, brooklyn_means, r_glm_poisson_ys)
bkfort = fortify(bklyn_tracts.shp, region="GEOID10")
pdata = left_join(bkfort, df1, by=c("id" = "bklyn_tractIDs"))

events_plot = ggplot() + geom_polygon(data=pdata, aes(x=long, y=lat, group=group, fill=bklyn_events)) +  coord_map() + coord_fixed() + scale_fill_gradientn(limits=c(0, 30), colors=blues9, oob=scales::squish, guide=guide_legend(title="")) + theme(axis.text.x=element_blank(), axis.text.y=element_blank(), axis.title.x=element_blank(), axis.title.y=element_blank()) + ggtitle("Events")

pop_plot = ggplot() + geom_polygon(data=pdata, aes(x=long, y=lat, group=group, fill=bklyn_pop)) + coord_map() + coord_fixed()  + scale_fill_gradientn(limits=c(0, 30), colors=blues9, oob=scales::squish, guide=guide_legend(title="")) + theme(axis.text.x=element_blank(), axis.text.y=element_blank(), axis.title.x=element_blank(), axis.title.y=element_blank()) + ggtitle("Population (rescaled)")

bym_plot = ggplot() + geom_polygon(data=pdata, aes(x=long, y=lat, group=group, fill=brooklyn_means)) + coord_map() + coord_fixed()  + scale_fill_gradientn(limits=c(0, 30), colors=blues9, oob=scales::squish, guide=guide_legend(title="")) + theme(axis.text.x=element_blank(), axis.text.y=element_blank(), axis.title.x=element_blank(), axis.title.y=element_blank()) + ggtitle("BYM fit - spatial smoothing")

poisson_plot = ggplot() + geom_polygon(data=pdata, aes(x=long, y=lat, group=group, fill=r_glm_poisson_ys)) + coord_map() + coord_fixed()  + scale_fill_gradientn(limits=c(0, 30), colors=blues9, oob=scales::squish, guide=guide_legend(title="")) + theme(axis.text.x=element_blank(), axis.text.y=element_blank(), axis.title.x=element_blank(), axis.title.y=element_blank()) + ggtitle("Pois-GLM fit - no smoothing")
```

```{r plot-brooklyn, echo = TRUE, comment = NA, message = FALSE}
grid.arrange(events_plot, pop_plot, bym_plot, poisson_plot, ncol=2)
```

### Discussion

The models developed in this case study allow for analysis of relatively large datasets;
for the last model in this case study, given a dataset with 709 areal units,
the fitted model had approximately 1400 parameters.

The scalability of IAR models derives from the nature of the covariance matrix
and the fact that Stan computes proportional densities, so that
constant terms drop out.  Furthermore, because spatial adjacency matrices are sparse
for large numbers of areal units $N$, efficient representations of the contents of these
matrices provides substantial speedup over trying to compute using a full $N \times N$ matrix.

### Acknowledgements

Many thanks to Daniel Simpson, Rob Trangucci, Imad Ali, Andrew Gelman, and Bob Carpenter for all their help and encouragement.


### References


#### Literature

* Banerjee, Sudipto, Alan E. Gelfand, and Bradley P. Carlin. "Hierarchical modeling and analysis for spatial data." (2003).

* Besag, Julian. "Spatial interaction and the statistical analysis of lattice systems." Journal of the Royal Statistical Society. Series B (Methodological) (1974): 192-236.

* Besag, Julian, and Charles Kooperberg. "On conditional and intrinsic autoregression." Biometrika (1995): 733-746.

* Besag, J., J. York, and A. Mollie. "Bayesian image restoration with two applications in spatial statistics (with discussion) Ann Inst Stat Math. 1991; 43: 1–59. doi: 10.1007." BF00116466.[Cross Ref].

* Haran, Murali. "Gaussian random field models for spatial data." Handbook of Markov Chain Monte Carlo (2011): 449-478.

#### R Packages

* Statistics: [RStan](http://mc-stan.org/rstan.html)

* Plots: [ggplot2](http://ggplot2.org), [dplyr](https://cran.r-project.org/web/packages/dplyr), [gridExtra](https://cran.r-project.org/web/packages/gridExtra)

* Spatial Data: [maptools](https://cran.r-project.org/web/packages/maptools),
[spdep](https://cran.r-project.org/web/packages/spdep),
[gpclib](https://cran.r-project.org/web/packages/gpclib)

### Licenses

<small>
**Code:** Copyright (2017) Columbia University.  Released under the
[BSD 3-clause license](https://opensource.org/licenses/BSD-3-Clause).
</small>

<small>
**Text:** Copyright (2017) Mitzi Morris.  Released under the
the [CC BY-NC 4.0
license](https://creativecommons.org/licenses/by-nc/4.0/).
</small>