---
title: "Spatial Models in Stan: Conditional and Intrinsic Auto-Regressive Models for Areal Data"
author: "Mitzi Morris"
output: 
  html_document:
    keep_md: false
---

When data has a spatio-temporal structure and when observations from
neighboring regions exhibit higher correlation than observations between
distant regions,  this correlation can be accounted for using the class
of spatial models called "CAR" models (Conditional Auto-Regressive)
introduced by Besag and collaborators.
This case study covers how to efficiently code these models in Stan.

## Spatial Data as a Gaussian Random Field Model

The following math and its notation is taken from
"Gaussian Random Field Models for Spatial Data" by Murali Haran,
which is Chapter 18 of the "Handbook of Markov Chain Monte Carlo".

Besag (1974) shows that by encoding the neighbor relations
between spatial regions as a lattice, results from
the physics of lattice systems of particles
and the Hammersly-Clifford theorem provide an equivalence between
a local specification of the conditional distribution of each particle
given its neighboring particles and the global specification
of the joint distribution of all particles.
This specification of the joint distribution via the local specification
of the conditional distributions of the individual variables
is a Markov random field specification.

Therefore, given a set of observations taken at $n$ different subregions
of a region with a number of dimensions $D$
(for spatio-temporal data, the number of dimensions is usually between 1 and 4,
i.e., 1-3 spatial dimensions and 1 time dimension),
spatial interactions between regions $n_i$ and $n_j$ can be modelled conditionally
as a spatial random variable $\mathbf{w}$ as follows:

- Let $w_{-i}$ denote the $n$-length vector $w$ excluding $w_i$.

- We model each $w_i$ in terms of its full conditional distribution which is
its distribution given the remaining random variables, $\mathbf{w}_{-i}$:
$$ w_i \,\vert\, \mathbf{w}_{-i}, \Theta \sim N
\left( \sum_{j=0}^n {c}_{ij} w_j,\kappa_i^{-1} \right),
i= 1,\ldots,n $$
where $c_{ij}$ describes the neighborhood structure such that  $c_{ij}$ is nonzero
only if $i$ and $j$ are neighbors and $\kappa$ is the precision (inverse variance) parameter.

### CAR Models

The neighborhood structure of the $\kappa$ and $c_{ij}$ elements can be stored
in an $n \times n$ matrix $Q$ where the diagonal elements
represent each of the $n$ subregions with value $\kappa_i$
and the off-diagonal elements contain $-\kappa_i c_{ij}$ 
if subregions $i$ and $j$ are adjacent and 0 otherwise.
Usually a common precision parameter $\tau$, is assumed,
where ${\kappa}_i = \tau$ for all $i$.

When the matrix $Q$ is symmetric and positive definite, this specifies a valid joint distribution,
$$ w \,\vert\, \Theta \sim  N(0, Q^{-1}) $$
with $\Theta$ the vector of the precision parameters.
This provides a proper prior for a CAR model.
However evaluation of $w$ requires computing the covariance matrix $Q^{-1}$,
which is computationally expensive for large values of $n$.

See the Stan case study 
[Exact sparse CAR models in Stan](http://mc-stan.org/documentation/case-studies/mbjoseph-CARStan.html),
for further discussion of this model.

### IAR Models

Intrisic Auto-Regressive models are 
intrinsic Gaussian Markov random fields,
(see Besag and Kooperberg 1995).
They are a subclass of CAR models which have an improper prior.
Spatial models which use this improper prior are most correctly called IAR models,
although often they are just called CAR models.

For this class of models, Q is a positive _semidefinite_ matrix.
The off-diagonals of Q are  $-{\tau} c_{ij}$
and the value of the $i^{th}$ diagonal element is $\tau \sum_{j} c_{ij}$.
This intrinsic GMRF model corresponds to the following conditional specification:

$$ f(w_j \vert \mathbf{w}_{-i}, \tau) \sim N \left(
\frac{ \sum_{j \in N(i)}^n w_j} {n},
\frac{1}{n_i,\tau} \right) $$

The individual spatial random variable $w_j$ for region $n_j$ with neighbors $N(i)$
is normally distributed with a mean equal to the average of its neighbors.
The variance decreases as the number of neighbors increases.

Although this is an improper prior, given data, this results in a proper posterior density.
Computing this density in Stan is computationally tractable due to the fact that
Stan is computing proportional densities, allowing constant terms to drop out;
among these is the term which requires computation of the determinant of this matrix.
The density of the random variable $\mathbf{w}$ is:

$$ f (\mathbf{w} \vert \Theta) \propto \tau^{(N−1)/2}  \exp(−\mathbf{w}^TQ(\tau)\mathbf{w}) $$

When the neighborhood graph contains disconnected subsets, term $\tau^{(N−1)/2}$
must be changed to $\tau^{(N−k)/2}$ where $k$ is the number of distinct subsets.

_NOTE:  this is only valid when all the regions in the adjacency matrix have at least 1 neighbor,
that is, for all subregions $n_i$, the set $j \sim i$ is non-empty._

An efficient representation of a sparse adjacency matrix for a set of $N$ subregions
is:

- an array of length $N$ which contains the number of neighbors for each region

- an $L \times 3$ array where $L$ is the number of the non-zero off-diagonal elements of the adjacency matrix.
The first two columns provide the $i,j$ region ids respectively, and the 3rd column contains
the value $c_{ij}$, (the weight contributed by this neighbor).
In the case where this value is always `-1`, the 3rd column can be omitted altogether.

The following Stan program fragments shows how to compute this in Stan.
The sparse adjacency matrix is passed in as data:
```
data {
  int<lower=1> diag_weights[N_regions];  // weights == num_neighbors
  int N_links; // number of non-zero entries in adj matrix
  int<lower=1> off_diag_coords[N_links,2]; // ij coords of off-diag entries
}
```
The IAR spatial component parameters are:
```
parameters {
  vector[N_regions] h;  // individual-level spatial effect (IAR)
  real<lower=0> tau;  // precision param
}
```

Stan computes the log density, i.e.,
$\frac{(N−1)}{2}log(\tau) + −\mathbf{w}^TQ(\tau)\mathbf{w}$.
The diagonal and off-diagonal elements are stored separately,
and therefore computed separately.
Because all off-diagonal values in the adjacency matrix are -1,
we specify this value directly as part of the model:
```
transformed parameters {
  real neg_tau_div_2 = -tau / 2;
}
model {
  real off_diag_weight = -1.0;
  ...
  tau ~ normal(0, 5);
  for (i in 1:N_tracts) { // diagonals
    target += neg_tau_div_2 * square(h[i]) * diag_weights[i];
  }
  for (j in 1:N_links) {   // off-diagonals
    target += neg_tau_div_2 *
    h[off_diag_coords[j,1]] * h[off_diag_coords[j,2]] * off_diag_weight;
  }
  target += ((N_tracts - 1) / 2.0) * log(tau);
  ...
}
```

## Spatial Models for Areal Count Data:  Poisson-CAR and Friends

Adding the CAR (or IAR) spatially structured error term to a multi-level GLM
as a random-effects multi-variate Gaussian 
provides spatial smoothing of the resulting estimates.
For count data, e.g. incidents of disease or accidents,
Poisson CAR models are available via R, BUGS, and JAGS
as well as INLA (Integrated Nested Laplace Approximation)
which is a fast alternative to MCMC,
(INLA trades speed and scalability for accuracy,
per the "no free lunch" principle).

For this case study, we show how to
create the Stan equivalent a Poisson GLM with an IAR component
using a subset of the dataset from
[Small-area spatiotemporal analysis of pedestrian and bicyclist injuries in New York City](https://www.ncbi.nlm.nih.gov/pubmed/25643104).

All models and data files are available in the Stan example-models GitHub repo for Stan case studies:
[car-iar-poisson](https://github.com/stan-dev/example-models/tree/master/knitr/car-iar-poisson).

### Preliminaries

All commands are run from working directory containing the data and models.
The following data, libraries, options needed for R are:

```{r loadstuff, comment = NA, message = FALSE, warning = FALSE}
library(gpclib);
library(maptools);  
library(spdep);
gpclibPermit()

library(dplyr) 
library(ggplot2) 
library(gridExtra)

library(rstan)  
rstan_options(auto_write = TRUE) 
options(mc.cores = parallel::detectCores())
```

### Data

Sample data is in file `nyc_ped_subset.data.R` which contains
census tract id, the count of events per tract (`y`),
and the population per tract (`x`).
Spatial information is in a set of files in directory `nycTracts10`.

```{r nyc-data-xy-plot, echo = FALSE, comment = NA, message = FALSE, warning = FALSE}
source("nyc_ped_subset.data.R");
y = events_all_2001[all_tractIDs %in% bklyn_tractIDs];
x = pop_adj_2001[all_tractIDs %in% bklyn_tractIDs];

plot(x,y,xlab="population (scaled)",ylab="observed events");
```

### Stan 101:  Poisson GLM

Use R to fit a Poisson model to NYC ped tract data for 2001.

```{r glm-poisson, echo = TRUE, comment = NA, message = FALSE, warning = FALSE}
source("nyc_ped_subset.data.R");
y = events_all_2001[all_tractIDs %in% bklyn_tractIDs];
x = pop_adj_2001[all_tractIDs %in% bklyn_tractIDs];

r_glm_poisson = glm(y~x,family="poisson")
print(r_glm_poisson)
```

The Stan equivalent of `glm(y~x,family="poisson")` for `N` observations worth of data
is in file "m1_poisson.stan":

```
data {
  int<lower=0> N;
  int<lower=0> y[N];
  vector[N] x;
}
parameters {
  real beta_1;
  real beta_2;
}
model {
  y ~ poisson_log(beta_1 + beta_2 * x);
}
generated quantities {
  vector[N] mu = exp(beta_1 + beta_2 * x);
}
```
We fit this model using the NUTS sampler (the default sampler for Rstan):

```{r rstan-poisson-fit, echo = TRUE, comment = NA, message = FALSE}
source("nyc_ped_subset.data.R");
y = events_all_2001[all_tractIDs %in% bklyn_tractIDs];
x = pop_adj_2001[all_tractIDs %in% bklyn_tractIDs];
N = length(bklyn_tractIDs);

m1_fit = stan("m1_poisson.stan",data=list(N,y,x))
print(m1_fit,pars=c("beta_1","beta_2"),probs=c(0.025,0.5,0.975))
traceplot(m1_fit,pars=c("lp__","beta_1","beta_2"))
```

The Stan model fits the data and produces the estimates for the
intercept (`beta_1`) and slope (`beta_2`) are the same as the R GLM estimates.
Out of a total of 4000 draws during sampling (4 chains, 1000 draws each),
the `Rhat` statistic is 1.00, indicating that the chains have converged and
the number of effective samples for the slope `alpha` and intercept `beta` are OK.
The traceplot shows good mixing of the chains.
(Rhat should be very close to 1.00 - values below 0.9 or above 1.1 are indicative of a problem).

### Adding the IAR component

We add the IAR component to the Poisson model using data structures
and programming techniques shown above to the model shown above,
see file `m4_poisson_IAR.stan`

In order to run this on the NYC ped data, we restrict our attention to Brooklyn tracts only.
The spatial information for the census tracts is obtained via the R `maptools` and `spdep` packages.
For the Stan program, we need to create inputs:  `N_tracts`, `x`, `y`, `N_links`, `diag_weights`, `off_diag_coords`.

First we create an `nb` object which is a list of all neighbors for each census tract.
Each list entry is itself a list containing the relative index of the neighboring regions.
The following R helper function takes an `nb` object which lists the neighbors for each census tract
as a list of lists and munges this into a sparse array structure required for the
input `off_diag_coords` in the above Stan program:

```
get_nb_off_diags = function(x) {
     n_tracts = length(x);
     n_links = sum(card(x));
     idx_i = vector(mode="numeric",length=n_links);
     idx_j = vector(mode="numeric",length=n_links);
     idx = 1;
     for (i in 1:n_tracts) {
        if (x[[i]][1] != 0) {
           for (j in 1:length(x[[i]])) {
              idx_i[idx] = i;
              idx_j[idx] = x[[i]][j];
              idx = idx + 1;
           }
        }
     }
     return(cbind(idx_i,idx_j));
}
```

The following sequence of R commands creates these inputs
for the Brooklyn subset of the data:

```{r munge-bk, echo = TRUE, comment = NA, message = FALSE, warning = FALSE}
# census tracts spatial info
nyc_all_tracts.shp<-readShapePoly("nycTracts10/nycTracts10");
# restrict to tracts in study, reorder
bklyn_tracts <- nyc_all_tracts.shp$GEOID10 %in% bklyn_tractIDs;
bklyn_tracts.shp <- nyc_all_tracts.shp[bklyn_tracts,]
bklyn_tracts.shp <- bklyn_tracts.shp[order(bklyn_tracts.shp$GEOID10),]
nb_bk = poly2nb(bklyn_tracts.shp);

# munge nb object into Stan model inputs
source("munge_data_helper.R") 
N_tracts = length(nb_bk);
diag_weights = card(nb_bk);  
N_links = sum(card(nb_bk));  
off_diag_coords = get_nb_off_diags(nb_bk);
```

Now let's try to fit a Poisson model which has a spatial effects component `h`
in addition to the Poisson GLM coefficients `beta_1` and `beta_2`.

```{r fit-poisson-car, echo = TRUE, comment = NA, message = FALSE}
m4_fit = stan("m4_poisson_IAR.stan",data=list(N_tracts,y,x,diag_weights,N_links,off_diag_coords), iter = 5000, warmup = 4000);
print(m4_fit,pars=c("lp__", "beta_1", "beta_2", "tau","h[1]","h[2]","h[3]","h[4]","h[5]","h[6]"),probs=c(0.025,0.5,0.975));
traceplot(m4_fit,pars=c("lp__", "beta_1", "beta_2","tau","h[1]","h[2]"));
```

This model fails to fit - it cannot estimate both the group-level intercept `beta_1`
and the spatial component `h` - these provide a textbook example of non-identifiability.
(Note that adding individual-level random effects component, centered or un-centered,
doesn't produce this problem, see models `m2_poisson_re.stan` and `m3_poisson_re_nc.stan`
included in the GitHub repo for this case study.)

Dropping the group-level intercept `beta_1` allows the model to fit.
In file `m4_poisson_IAR_v2.stan` we remove the parameter `beta_1`, so
the Poisson GLM is fit using just the rate and the spatial component `h`:
```
data {
  int<lower=1> N_tracts; // number of census tracts in study
  int<lower=0> y[N_tracts];
  vector[N_tracts] x;
  int<lower=1> diag_weights[N_tracts]; // weights == num_neighbors
  int<lower=1> N_links; // number of non-zero entries in adj matrix
  int<lower=1> off_diag_coords[N_links,2]; // ij coords of off-diag entries
}
parameters {
  real beta_2;
  vector[N_tracts] h;  // individual-level spatial effect (IAR)
  real<lower=0> tau;  // precision param
}
transformed parameters {
  real neg_tau_div_2 = -tau * 0.5;
}
model {
  real off_diag_weight = -1.0;
  y ~ poisson_log(beta_2 * x + h);
  beta_2 ~ normal(0, 2.5);
  tau ~ normal(0, 5);
  for (i in 1:N_tracts) { // diagonals
    target += neg_tau_div_2 * square(h[i]) * diag_weights[i];
  }
  for (j in 1:N_links) {   // off-diagonals
    target += neg_tau_div_2
              * h[off_diag_coords[j,1]] * h[off_diag_coords[j,2]] * off_diag_weight;
  }
  target += ((N_tracts - 1) * 0.5) * log(tau);
}
generated quantities {
  vector[N_tracts] mu = exp(beta_2 * x + h);
}
```

The model diagnostics and trace plots indicate that the model fits the data, given proper amounts of warm-up iterations.

```{r fit-poisson-car-v2, echo = TRUE, comment = NA, message = FALSE}
m4_v2_fit = stan("m4_poisson_IAR_v2.stan",data=list(N_tracts,y,x,diag_weights,N_links,off_diag_coords), iter = 10000, warmup = 9000);
print(m4_v2_fit,pars=c("lp__","beta_2", "tau","h[1]","h[2]","h[3]","h[4]","h[5]","h[6]"),probs=c(0.025,0.5,0.975));
traceplot(m4_v2_fit,pars=c("lp__","beta_2","tau","h[1]","h[2]","h[3]"));
```

This model fits the data, although the group-level and some of the 
some of the individual-level spatial effects parameters (`h[1]` et. al.)
have low numbers of effective samples (EFF).

We add a individual-level random-effects component.
There are 709 census tracts in the Brooklyn dataset we're using, so this brings the number of
parameters to estimate to 1421.
Model fitting and sampling takes noticeably longer.

```{r fit-poisson-car-re, echo = TRUE, comment = NA, message = FALSE}
m5_fit = stan("m5_poisson_IAR_re_nc.stan",data=list(N_tracts,y,x,diag_weights,N_links,off_diag_coords), iter = 10000, warmup = 9000);
print(m5_fit,pars=c("lp__","beta_2", "tau","sigma","h[1]","h[2]","h[3]","h[4]","h[5]","h[6]"),probs=c(0.025,0.5,0.975));
traceplot(m5_fit,pars=c("lp__", "beta_2","tau","sigma","h[1]", "h[2]", "re_nc[1]", "re_nc[2]"));
```

We `ggplot2` and friends to plot spatial relationships to plot
the Brooklyn census tracts colored by event counts, scaled population,
and the generated quantity "Mu" from models `m4_poisson_IAR_v2.stan`
and `m4_poisson_IAR_re_nc.stan`, where Mu is the number of
events based on the data and estimated model pameters.
Note that some Brooklyn census tracts are not in the study.

```{r plot-poisson-car, echo = TRUE, comment = NA, message = FALSE}
# get estimates from fits for m4 and m5
mu_names = paste("mu[",c(1:709),"]")
mu4_samples = as.data.frame(m4_v2_fit,pars=mu_names)
mu4_means = apply(mu4_samples,2,mean)
mu5_samples = as.data.frame(m5_fit,pars=mu_names)
mu5_means = apply(mu5_samples,2,mean)

# prepare dfs for ggplot2
bklyn_events = events_all_2001[bklyn_tracts]
bklyn_pop = pop_adj_2001[bklyn_tracts]
df1 = data.frame(bklyn_tractIDs,bklyn_events,bklyn_pop,mu4_means,mu5_means)
bkfort = fortify(bklyn_tracts.shp, region="GEOID10")
pdata = left_join(bkfort,df1,by=c("id" = "bklyn_tractIDs"))

events_plot = ggplot() + geom_polygon(data=pdata, aes(x=long, y=lat, group=group,fill=bklyn_events)) +  coord_map() + coord_fixed() + scale_fill_gradientn(limits=c(0,30), colors=blues9, oob=scales::squish,guide=guide_legend(title="")) + theme(axis.text.x=element_blank(),axis.text.y=element_blank(),axis.title.x=element_blank(),axis.title.y=element_blank()) + ggtitle("Events")

pop_plot = ggplot() + geom_polygon(data=pdata, aes(x=long, y=lat, group=group, fill=bklyn_pop)) + coord_map() + coord_fixed()  + scale_fill_gradientn(limits=c(0,30), colors=blues9, oob=scales::squish,guide=guide_legend(title="")) + theme(axis.text.x=element_blank(),axis.text.y=element_blank(),axis.title.x=element_blank(),axis.title.y=element_blank()) + ggtitle("Population (rescaled)")

m4_mu_plot = ggplot() + geom_polygon(data=pdata, aes(x=long, y=lat, group=group, fill=mu4_means)) + coord_map() + coord_fixed()  + scale_fill_gradientn(limits=c(0,30), colors=blues9, oob=scales::squish,guide=guide_legend(title="")) + theme(axis.text.x=element_blank(),axis.text.y=element_blank(),axis.title.x=element_blank(),axis.title.y=element_blank()) + ggtitle("Mu (estimated events), model m4")

m5_mu_plot = ggplot() + geom_polygon(data=pdata, aes(x=long, y=lat, group=group, fill=mu5_means)) + coord_map() + coord_fixed()  + scale_fill_gradientn(limits=c(0,30), colors=blues9, oob=scales::squish,guide=guide_legend(title="")) + theme(axis.text.x=element_blank(),axis.text.y=element_blank(),axis.title.x=element_blank(),axis.title.y=element_blank()) + ggtitle("Mu (estimated events), model m5")

grid.arrange(events_plot, pop_plot, m4_mu_plot, m5_mu_plot, ncol=2)
```

### Discussion

The models developed in this case study allow for analysis of relatively large datasets;
for the last model in this case study, given a dataset with 709 subregions,
the fitted model had approximately 1400 parameters.
Running this on a 2-core MacBook Pro took approximately 5 minutes.
Due to the improper prior, IAR models are not generative;
therefore we cannot create simulated data with which to develop and test this model.

The scalability of IAR models, as noted at the outset, derives from the nature of the
covariance matrix Q and the fact that Stan computes proportional densities, so that
constant terms drop out.  Furthermore, because spatial adjacency matrices are sparse
for large numbers of subregions $N$, efficient representations of the contents of these
matrices provides substantial speedup over trying to compute using a full $N \times N$ matrix.


### References

#### Papers

* Besag, Julian. "Spatial interaction and the statistical analysis of lattice systems." Journal of the Royal Statistical Society. Series B (Methodological) (1974): 192-236.

* Besag, Julian, and Charles Kooperberg. "On conditional and intrinsic autoregression." Biometrika (1995): 733-746.

* Haran, Murali. "Gaussian random field models for spatial data." Handbook of Markov Chain Monte Carlo (2011): 449-478.

#### R Packages

* Statistics: [RStan](http://mc-stan.org/rstan.html)

* Plots: [ggplot2](http://ggplot2.org), [dplyr](https://cran.r-project.org/web/packages/dplyr)

* Spatial Data: [maptools](https://cran.r-project.org/web/packages/maptools),
[spdep](https://cran.r-project.org/web/packages/spdep),
[gpclib](https://cran.r-project.org/web/packages/gpclib)

### Licenses

<small>
**Code:** Copyright (2017) Columbia University.  Released under the
[BSD 3-clause license](https://opensource.org/licenses/BSD-3-Clause).
</small>

<small>
**Text:** Copyright (2017) Mitzi Morris.  Released under the
the [CC BY-NC 4.0
license](https://creativecommons.org/licenses/by-nc/4.0/).
</small>