---
title: "Validating Bayesian Samplers with Simulated Data"
author: "Bob Carpenter"
date: "30 May 2017"
output:
  html_document:
    fig_caption: yes
    fig_height: 2
    fig_width: 3
    theme: cerulean
    toc: yes
    toc_depth: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract

Clear specification and rigorous testing is the foundation of robust
software development.  This case study demonstrates a general approach
to testing Bayesian sampling algorithms due to Cook, Gelman, and Rubin
and oultined in

* Cook, Samantha R., Andrew Gelman, and Donald
  B. Rubin. 2006. [Validation of software for Bayesian models using
  posterior
  quantiles](http://stat.columbia.edu/~gelman/research/published/Cook_Software_Validation.pdf).  *Journal of Computational and Graphical Statistics* 15(3):675--692.

The basic idea is to repeatedly simulate data from a specific Bayesian
model; this is done by simulating parameters from the prior and then
simulating observed data from the sampling distribution.  If the
software is working correctly, the posteriors it produces will be
properly calibrated under repeated simulation. For example, roughly
50% of the 50% intervals will contain the simulated parameter value.
Cook et al. convert this into a statistical test, the calculation of
which we demonstrate in R for Bayesian models implemented in Stan.


## Outline

The first section summarizes Cook et al.'s clever proof that Bayesian
posterior inference is well calibrated in general.  The second section
shows how they use this result to generate a test statistic to
validate posterior inferences for the model.  The remainder of the
case study illustrates the application of their method to Stan
programs.

# Bayesian Model Calibration


### Bayesian model specification

We will assume that a Bayesian model is provided in the usual way as a
prior and sampling distribution.  Specifically we assume that

* the prior is given as a probability function $p(\theta)$
  over model parameters $\theta \in \mathbb{R}^K$, and

* the sampling idstirbution is given as a probability function $p(y
  \mid \theta)$ over a vector $y \in \mathbb{R}^N$ of data given the
  model parameters $\theta$.

<div style="width:35%; float:right; padding:1em 1em 1em 1em">
![<small>Figure 1. <i>The steps involved for one replication of Cook
et al.'s validation process.  Image from (Cook et al. 2006, figure
2)</i></small>.](cook-gelman-rubin-fig2.png)
</div>

### Simulating parameters and data

Suppose we have a secondary program with which we

* simulate parameters $\theta^{(0)}$ according to the prior
  distribution $p(\theta)$, and

* simulate data $y^{(0)}$ according to the sampling distribution
  $p(y \mid \theta^{(0)})$.

The chain rule shows that this is also

* a draw $\left( y^{(0)}, \theta^{(0)} \right)$ from the joint distribution
  $p(y, \theta) \propto p(y \mid \theta) \ p(\theta)$.

From Bayes's rule, we see that we now have

* a draw $\theta^{(0)}$ from the posterior distribution $p(\theta \mid y^{(0)}) \propto p(y^{(0)}, \theta) \, = \, p(y^{(0)} \mid \theta) \, p(\theta)$.


### MCMC calculation of posterior quantiles

Now suppose we have a proposed method for sampling, the goal of which is to

* draw $\theta^{(1)}, \ldots, \theta^{(M)}$ from the posterior $p(\theta \mid y^{(0)})$

Then we can estimate posterior quantiles for the components of
$\theta^{(0)}$ based on the draws $\theta^{(1)}, \ldots \theta^{(M)}$, as


\[
{\hat{q}}{\mbox{}_k}
\ \approx \
\frac{1}{M} \sum_{m = 1}^M \mathrm{I}[ \theta^{(m)}_k < \theta^{(0)}_k ].
\]


### Distribution of quantiles

Because $\theta^{(0)}$ is just a random draw from the posterior, any
of its components is equally likely to land in any quantile in $(0,
1)$.  The approximation to the true quantile improves with $M$, so
that as $M \rightarrow \infty$,

\[
\hat{q}_k \ \sim \ \mathsf{Uniform}(0, 1).
\]

Cook et al. provide a slightly more rigorous proof of this (their Thereom 1).
For practical applications, convergence within a finite $M$ will
require geometric ergodicity of the sampler for the model, a property
which will be tested along with other aspects of the sampler and model
implementation.

### Ramification for Bayesian posterior calibration

As a consequence of the quantiles being uniform, we automatically get calibrated posterior coverage.  Specifically, if we simulate $\theta^{(0)}$ from the prior and $y^{(0)}$ from the sampling distribution conditioned on $\theta^{(0)}$, the the posterior $p(\theta \mid y^{(0)})$ should have $p\%$ intervals that can be expected to contain the simulated parameter value $p\%$ of the time.


# Test Statistic

Cook et al. develop a test statistic for Bayesian posterior samplers
based on the uniformity of the posterior quantile estimates $\hat{q}_k$.

### Transforming quantiles to z-scores

In general, if $u \sim \mathsf{Uniform}(0, 1)$, then

\[
z(u) = \Phi^{-1}(u) \sim \mathsf{Normal}(0, 1),
\]

where $\Phi$ is the standard normal cumulative distribution function,

\[
\Phi(v)
\ = \
\int_{-\infty}^v
\mathsf{Normal}(w \mid 0, 1) \, \mathrm{d}w,
\]
and $\Phi^{-1}$ is its inverse (i.e., the function that maps a
quantile value in $(0, 1)$ to the value in the standard normal
distribution with that quantile).

### Sum of squared z-scores is $\chi^2$

Cook et al. define a test statitisic that is equal to the sum of the
squared z-scores for $I$ replicated tests, each of which generates a
new $\theta^{(0)}$ and $y^{(0)}$ and samples a sequence $\theta^{(1)},
\ldots, \theta^{(M)}$ from the corresponding posterior $p(\theta \mid y^{(0)})$, to generate a posterior quantile estimate
$\hat{q}^{(i)}_k$ for the $k$-th component of the posterior,

\[
X^2_{\theta_k} = \sum_{i=1}^I \left( \Phi^{-1}(\hat{q}^{(i)}_k) \right)^2.
\]

Because each of the terms is independently drawn as standard normal,
the sum of their squares, $X^2_{\theta_k}$, will have a standard $\chi^2$
distribution with $I$ degrees of freedom, i.e.,

\[
X^2_{\theta_k} \sim \mathsf{ChiSquare}(I, 0, 1).
\]


### Detecting problems with posterior sampling

As Cook et al. note, the $X^2_{\theta_k}$ statistic can be used to calculate a $p$-value for the null hypotehsis that $X^2_{\theta_k}$ has a $\chi^2$ distribution with $I$ degrees of freedom.  Unexpectedly small $p$-values indicate a mismatch between the process used to generate the simulated data and parameters $y^{(0)}, \theta^{(0)}$, and the process used to generate posterior draws $\theta^{(1)}, \ldots, \theta^{(M)}$ conditioned on $y^{(0)}$.

### Adjusting false alarm rate for multiple comparisons

Cook et al. apply a so-called Bonferronni correction to control the
false alarm rate under multiple comparisons.  Specifically, they
multiply each $p$-value by $K$, because $K$ components are being
tested.


# Example 1: Hierarchical Normal

We will illustrate Cook et al.'s approach with a variant of their
simple varying-intercepts hierarchical model that more closely follows
current thinking about priors.

### Varying-intercepts hierarchical normal model

The data consists of real-valued observations $y_{i, j}$ organized
into $j \in 1{:}J$ groups with $N_j$ observations per group.  Given
the parameters, the sampling distribution is

* $y_{i, j} \sim \mathsf{Normal}(\alpha_j, \sigma)$ for $j \in 1{:}J$
  and $i \in 1{:}N_j$

with the intercept for group $j$ being drawn from the hierarchical
prior

* $\alpha_j \sim \mathsf{Normal}(\mu, \tau)$ for $j \in 1{:}J$.

The prior for overall observation noise scale $\sigma > 0$ is

* $\sigma \sim \mathsf{HalfNormal}(0, 5)$

and the hyperpriors for the prior location $\mu$ and scale $\tau > 0$ are

* $\mu \sim \mathrm{Normal}(5, 5)$, and

* $\tau \sim \mathsf{HalfNormal}(0, 5)$.

### Simulating data from the model

We will simulate the data using R.  We could simulate the data using
Stan, but there is a benefit to using two independent implementations
in testing.

First, we set the seed for replicability across runs of this knitr
file and then simulate the data.

```{r}
set.seed(4321);
J <- 6;
N <- c(33, 21, 22, 22, 24, 11);
mu <- rnorm(1, 5, 5);
tau <- abs(rnorm(1, 0, 5));
sigma <- abs(rnorm(1, 0, 5));
alpha <- rnorm(J, mu, tau);
y <- c();
begin <- 1;
for (j in 1:J) {
  end <- begin + N[j] - 1;
  y[begin:end] <- rnorm(N[j], alpha[j], sigma);
  begin <- end + 1;
}
```

This yields the following parameters

```{r}
print(sprintf("tau:   %4.1f", tau), quote=FALSE);
print(sprintf("mu:    %4.1f", mu), quote=FALSE);
print(sprintf("sigma: %4.1f", sigma), quote=FALSE);
```

The intercepts for the groups are

```{r}
print(sprintf("alpha[%1d]: %3.1f   ", 1:J, alpha), quote=FALSE);
```

Together, these yield the simulated observations,

```{r}
print(y, digits=1, quote=FALSE)
```

### Coding the model in Stan

The Stan program reflects the model definition printed above with two
caveats.  First, Stan does not provide a ragged array data structure,
so there is a fiddly calculation of the group index for each
observation in the transformed data block.  Second, the Stan program
uses the non-centered parameterization for the varying intercepts
<code>alpha</code> because we know there are not many observations for
each parameter.  The full Stan program is as follows.

```{r comment=NA, echo=FALSE, tidy=TRUE}
file_path <- "norm-hier-nonctr.stan";
lines <- readLines(file_path, encoding="ASCII");
for (n in 1:length(lines)) cat(lines[n],'\n');
```

The centered parameterization of <code>alpha</code> is how the model
was originally defined.  The centered parameterization would be coded
in Stan as follows.

```
parameters {
  vector[J] alpha;
  ...
model {
  alpha ~ normal(mu, tau);
```

The centered parameterization mixes very poorly when the group
observations do not tightly constrain the hierarchical parameters.
The resulting posterior will vary greatly as <code>tau</code> varies
and the <code>alpha</code> values depart from <code>mu</code>.

The non-centered parameterization is much better behaved in such
cases.  It uses the parameter <code>alpha_std</code> with a standard
normal distribution and recovers <code>alpha</code> as a transformed
parameter.

```
parameters {
  vector[J] alpha_std;
  ...
transformed parameters {
  vector[J] alpha = mu + alpha_std * tau;
}
model {
  alpha_std ~ normal(0, 1);
```

The trick is that the non-centered parameterization removes the
dependency in the prior between the low level parameters
<code>alpha</code> and the hyperprior parameters <code>mu</code> and
<code>tau</code>.


### Fitting the model

We can now fit the simulated model using Stan using the default
configuration of four chains with 1000 warmup and 1000 sampling
iterations each.
```{r }
library(rstan);
rstan_options(auto_write = TRUE)
options(mc.cores = 5);

fit <- stan("norm-hier-nonctr.stan",
            data = c("J", "N", "y"),
            control = list(stepsize = 0.001,
                           adapt_delta = 0.99),
            seed = 12345,
            refresh = 0, open_progress=FALSE);
```

and then print the results for the parameters of interest,

```{r}
print(fit, pars = c("mu", "tau", "sigma", "alpha"),
      probs = c(0.1, 0.9),
      digits = 2);
```

The fit exhibits no signs of failing to mix---all R-hat values near 1
and all effective sample sizes near the number of posterior
iterations.

Next, we estimate the quantiles of our original simulations,

```{r}
q_hat <- function(fit) {
  posterior_draws <- extract(fit);
  qh <- c();
  qh[1] <- mean(posterior_draws$mu > mu);
  qh[2] <- mean(posterior_draws$tau > tau);
  qh[3] <- mean(posterior_draws$sigma > sigma);
  for (j in 1:J)
    qh[3 + j] <- mean(posterior_draws$alpha[ , j] > alpha[j]);
  return(qh);
}
```

and provide the z-transform function for the fit,

```{r}
z_hat <- function(fit) {
  return(qnorm(q_hat(fit)));
}
```

Applied to the fit above, this yields the quantiles

```{r}
print(q_hat(fit), digits=2);
```

and the z-transformed quantiles

```{r}
print(z_hat(fit), digits=2);
```

### Replicating the simulation

Now we can replicate the entire experiment 100 times and plot
the results.

```{R}
q_hat_mu <- c();     q_hat_tau <- c();
q_hat_sigma <- c();  q_hat_alpha1 <- c();

z_hat_mu <- c();     z_hat_tau <- c();
z_hat_sigma <- c();  z_hat_alpha1 <- c();

I <- 4;
for (i in 1:I) {
  # simulate
  J <- 6;
  N <- c(33, 21, 22, 22, 24, 11);
  mu <- rnorm(1, 5, 2);
  tau <- abs(rnorm(1, 0, 3));
  sigma <- abs(rnorm(1, 0, 2.5));
  alpha <- rnorm(J, mu, tau);
  y <- c();
  begin <- 1;
  for (j in 1:J) {
    end <- begin + N[j] - 1;
    y[begin:end] <- rnorm(N[j], alpha[j], sigma);
    begin <- end + 1;
  }

  # fit
  fit <- stan("norm-hier-nonctr.stan",
              data = c("J", "N", "y"),
              control = list(stepsize = 0.001,
                             adapt_delta = 0.99),
              iter = 5000,
              refresh = 0, open_progress=FALSE);
  draws <- extract(fit);

  # need a bit of a prior here on the mean

  # q_hat
  q_hat_mu[i] <- mean(draws$mu > mu);
  q_hat_tau[i] <- mean(draws$tau > tau);
  q_hat_sigma[i] <- mean(draws$sigma > sigma);
  q_hat_alpha1[i] <- mean(draws$alpha[, 1] > alpha[1]);

  # z_hat
  z_hat_mu[i] <- qnorm(q_hat_mu[i]);
  z_hat_tau[i] <- qnorm(q_hat_tau[i]);
  z_hat_sigma[i] <- qnorm(q_hat_sigma[i]);
  z_hat_alpha1[i] <- qnorm(q_hat_alpha1[i]);
}
df_q_hat <- data.frame(q_hat_mu,    q_hat_tau,
                       q_hat_sigma, q_hat_alpha1);
df_z_hat <- data.frame(z_hat_mu,    z_hat_tau,
                       z_hat_sigma, z_hat_alpha1);
x_sq_mu = sum(z_hat_mu^2);
x_sq_tau = sum(z_hat_tau^2);
x_sq_sigma = sum(z_hat_sigma^2);
x_sq_alpha1 = sum(z_hat_alpha1^2);

p_mu = pchisq(x_sq_mu, I);
p_tau = pchisq(x_sq_tau, I);
p_sigma = pchisq(x_sq_sigma, I);
p_alpha1 = pchisq(x_sq_alpha1, I);
```


# Misspecified Models and Real Data

If it turns out that the data is *not* generated from the model's
generated process, that is, if the model is misspecified, then all
bets are off.  The posterior calibration results demonstrated in the
first section are all conditional on using the true model to generate
the data.




