---
title: "Rasch model with latent regression"
author: "Daniel C. Furr"
date: "`r gsub('(^.* )(0)([0-9], .*)', '\\1\\3', format(Sys.time(), '%B %d, %Y'))`"
output:
  html_document:
    toc: true
    number_sections: true
    fig_caption: true
    css: styles.css
bibliography: bibliography.bib
---

<!-- 
(Title:)  Rasch model with latent regression

Author:  Daniel C. Furr

Date:  2016

Abstract:  This case study documents a Stan model for the Rasch model with latent regression. The latent regression portion of the model may be restricted to an intercept only, yielding a standard Rasch model. A brief simulation indicates that the Stan model successfully recovers the generating parameters. An example using a grade 12 science assessment is provided.

Keywords:  education, IRT, Rasch model, 2PL. 
--> 

```{r, include = FALSE}
knitr::opts_chunk$set(tidy = TRUE, cache = TRUE)
```


# Model

## Overview

The Rasch model [@Rasch1960a] is an item response theory model for dichotomous items. The version presented includes a latent regression. However, the latent regression part of the model may be restricted to an intercept only, resulting in a regular Rasch model.

$$ 
\mathrm{logit} [ \Pr(y_{ij} = 1 | \theta_j, \beta_i) ] = 
  \theta_j - \beta_i
$$
$$
\theta_j \sim \mathrm{N}(w_j' \lambda, \sigma^2)
$$

Variables:

* $i = 1 \ldots I$ indexes items
* $j = 1 \ldots J$ indexes persons
* $y_{ij} \in \{ 0 \ldots m_i \}$ is the response of person $j$ to item $i$
* $w_{j}$ is the vector of covariates for person $j$

Parameters:

* $\beta_i$ is the difficulty for item $i$
* $\theta_j$ is the ability for person $j$
* $\lambda$ is the vector of latent regression parameters
* $\sigma^2$ is the variance for the ability distribution


## **Stan** program

The **Stan** program is nearly a direct translation of the notation above. The sole exception is a constraint placed on the item difficulties for model identification. This is done by creating `beta_free`, which is the vector of unconstrained item parameters, in the parameters block. Then in the transformed parameters block, `beta` is created to be identical to `beta_free` except for one additional element that is the constrained item difficulty. As a result of this constraint, the mean of `beta` will be zero.

```{r stan_code, tidy = TRUE, comment="", echo=FALSE}
cat(readLines("rasch_latent_reg.stan"), sep = "\n")
```


# Simulation

First, the necessary **R** packages are loaded.

```{r, message=FALSE, warning=FALSE, results="hide"}
# Load R packages
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
library(ggplot2)
```

The **R** code that follows simulates a dataset conforming to the model. The **Stan** model will be evaluated in terms of its ability to recover the generating values of the parameters when fit to this dataset.

```{r sim_data}
# Set paramters for the simulated data
I <- 20
J <- 500
sigma <- 1
lambda <- c(.5, .5, .5)
W <- cbind(1, rnorm(J, 0, 1), rnorm(J, 0, 1))
beta_free <- seq(from = -2, to = 2, length.out = I-1)

# Calculate or sample remaining paramters
theta <-  W %*% matrix(lambda) + rnorm(J, 0, sigma)
beta <- c(beta_free, -1 * sum(beta_free))

# Assemble data and simulate response
sim_data <- list(I = I,
                 J = J,
                 N = I*J,
                 ii = rep(1:I, times = J),
                 jj = rep(1:J, each = I))
eta <- theta[sim_data$jj] - beta[sim_data$ii]
sim_data$y <- as.numeric(boot::inv.logit(eta) > runif(sim_data$N))
sim_data$K <- ncol(W)
sim_data$W <- W
```

The simulated data consists of `r I` dichotomous items and `r J` persons. The latent regression includes an intercept and `r length(lambda) - 1` person-related covariates, which are standard normal variables. The simulated dataset is next fit with **Stan**.

```{r sim_fit, results='hide'}
# Fit model to simulated data
sim_fit <- stan(file = "rasch_latent_reg.stan", 
                data = sim_data, chains = 4, iter = 200)
```

Before interpreting the results, it is necessary to check that the chains have converged. **Stan** provides the $\hat{R}$ statistic for the model parameters and log posterior. These are provided in the following figure. All values for $\hat{R}$ should be less than 1.1.

```{r sim_converge, fig.cap="Convergence statistics ($\\hat{R}$) by parameter for the simulation. All values should be less than 1.1 to infer convergence."}
# Plot of convergence statistics
sim_monitor <- as.data.frame(monitor(sim_fit, print = FALSE))
sim_monitor$Parameter <- as.factor(gsub("\\[.*]", "", rownames(sim_monitor)))
ggplot(sim_monitor) +
  aes(x = Parameter, y = Rhat, color = Parameter) +
  geom_jitter(height = 0, width = .5, show.legend = FALSE) +
  ylab(expression(hat(italic(R))))
```

The **Stan** model is evaluated in terms of its ability to recover the generating values of the parameters. The R code below prepares a plot in which the points indicate the difference between the posterior means and generating values for the parameters of main interest. This difference is referred to as discrepancy. The lines indicate the 95% posterior intervals for the difference. Ideally, (nearly) all the 95% posterior intervals would include zero.

```{r sim_plot, fig.cap="Discrepancies between estimated and generating parameters. Points indicate the difference between the posterior means and generating values for a parameter, and horizontal lines indicate 95% posterior intervals for the difference. Most of the discrepancies are about zero, indicating that **Stan** successfully recovers the true parameters."}
# Make vector of wanted parameter names
wanted_pars <- c(paste0("beta[", 1:sim_data$I, "]"), 
                 paste0("lambda[", 1:sim_data$K, "]"), 
                 "sigma")

# Get estimated and generating values for wanted parameters
generating_values = c(beta, lambda, sigma)
sim_monitor <- monitor(sim_fit, probs = c(.025, .975), print = FALSE)
estimated_values <- sim_monitor[wanted_pars, c("mean", "2.5%", "97.5%")]

# Assesmble a data frame to pass to ggplot()
sim_df <- data.frame(parameter = factor(wanted_pars, rev(wanted_pars)),
                     row.names = NULL)
sim_df$middle <- estimated_values[,"mean"] - generating_values
sim_df$lower <- estimated_values[,"2.5%"] - generating_values
sim_df$upper <- estimated_values[,"97.5%"] - generating_values

# Plot the discrepancy
ggplot(sim_df) +
  aes(x = parameter, y = middle, ymin = lower, ymax = upper) +
  scale_x_discrete() +
  labs(y = "Discrepancy", x = NULL) +
       #title = "Discrepancies between estimated and generating parameters") +
  geom_abline(intercept = 0, slope = 0, color = "white") +
  geom_linerange() +
  geom_point(size = 2) +
  theme(panel.grid = element_blank()) +
  coord_flip()
```


# Example application

The example data [@testfact4] are from a grade 12 science assessment. 

```{r example_data, results='hide', warning=FALSE, message=FALSE, tidy = TRUE}
# Use data and scoring function from the mirt package
library(mirt)
sat <- key2binary(SAT12,
    key = c(1,4,5,2,3,1,2,1,3,1,2,4,2,1,5,3,4,4,1,4,3,3,4,1,3,5,1,3,1,5,4,5))
```

`r nrow(sat)` students responded to `r ncol(sat)` dichotomously scored items. Non-responses were scored as incorrect, so the data contain no missing values. The scored response matrix is converted to list form and fit with **Stan**. Because no person covariates are available, the covariate matrix `W` contains only an intercept term.

```{r example_fit, cache=TRUE, results='hide', warning=FALSE, message=FALSE, tidy = TRUE}
# Assemble data list and fit model
sat_data <- list(I = ncol(sat),
                 J = nrow(sat),
                 N = length(sat),
                 ii = rep(1:ncol(sat), each = nrow(sat)),
                 jj = rep(1:nrow(sat), times = ncol(sat)),
                 y = as.vector(sat),
                 K = 1,
                 W = matrix(1, nrow = nrow(sat), ncol = 1))
sat_fit <- stan(file = "rasch_latent_reg.stan", 
                data = sat_data, chains = 4, iter = 200)
```

As discussed above, convergence of the chains is assessed for every parameter, and also the log posterior density, using $\hat{R}$.

```{r example_converge, tidy=TRUE, fig.cap="Convergence statistics ($\\hat{R}$) by parameter for the example. All values should be less than 1.1 to infer convergence."}
# Plot of convergence statistics
sat_monitor <- as.data.frame(monitor(sat_fit, print = FALSE))
sat_monitor$Parameter <- as.factor(gsub("\\[.*]", "", rownames(sat_monitor)))
ggplot(sat_monitor) +
  aes(x = Parameter, y = Rhat, color = Parameter) +
  geom_jitter(height = 0, width = .5, show.legend = FALSE) +
  ylab(expression(hat(italic(R))))
```

Next we view summaries of the parameter posteriors.

```{r example_print, tidy = TRUE}
# View table of parameter posteriors
print(sat_fit, pars = c("beta", "lambda", "sigma"))
```


# References

<!-- This comment causes section to be numbered -->
